---
title: "GIS Code"
author: "Rebecca Irons (IRNREB001)"
date: '2022-04-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GIS Assignment

#### Section 1: Install and Load Packages 
#### Section 2: Tidy SAFRING Data
#### Section 3: Tidy Geographic Pentad Data
#### Section 4: Creating the Map 

This document provides step-by-step instructions to create a map of the localities of where 
malachite sunbirds have been ringed across South Africa, from east by KwaZulu-Natal to west. The data for this species is accessed from SAFRING using an API. For more details on the SAFRING database, see <https://safring.birdmap.africa/>. 

### Section 1

In this section, we will install and load the necessary packages to read in the data from the SAFRING database, tidy the data and use GIS tools to create a map. 

First, the workspace must be clean. 
```{r step 1}
rm(list = ls())
```

Install the necessary packages: 

- Tidyverse is required to manipulate the data: 
```{r step 2a, results='hide', warning=FALSE}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
```

- RCurl and rjson is required to access the data from SABAP2 using as API.  
```{r step 2b, results='hide', warning=FALSE}
install.packages("RCurl", repos = "http://cran.us.r-project.org")

install.packages("rjson", repos = "http://cran.us.r-project.org")
```

- sf is required to work with vector data to create a map
```{r step 2c, results='hide', warning=FALSE}

install.packages("sf", repos = "http://cran.us.r-project.org")

```


The packages from the tidyverse have to be loaded. These packages are:

```{r step 3a, results='hide', warning=FALSE, message=FALSE}

library(dplyr)

library(ggplot2)

library(broom)

```


The packages required for loading in the data from the SAFRING website are:
```{r step 3 b, results='hide',warning=FALSE, message=FALSE}

library(RCurl)

library(rjson)
```

The packages required for working with the vector data are: 
```{r step 3c, results='hide', warning=FALSE, message=FALSE}

library(sf)
```


Using an API, access the SAFRING database and load the data for your focal species. 

```{r step 4, results='hide'}
SPP = 751
ringing_data <- read.csv(paste('https://api.birdmap.africa/safring/species/records_list/',SPP,'?format=csv',sep=""))
```


### Section 2

The species data accessed from the SAFRING database need to be tidied first before we look at the spatial data. This is so that the dataset from which we are drawing the pentad data contains only observations that are sensical. These same steps have been followed in the data management assignment for tidying the SAFRING data. 

The ages that are recorded for the birds in the database must be numeric and the juveniles must be filtered out from the adults. We will only be working with adults. 

```{r step 5, results='hide'}

ringing_data$Age <- as.numeric(ringing_data$Age)

ringing_data <- filter(ringing_data, Age>3)
```


The metric data (tarsus length, mass etc) needs to be tidied. These data must be numeric.
```{r step 6, results='hide'}

ringing_data$Mass <- as.numeric(as.character(ringing_data$Mass))

ringing_data$Wing <- as.numeric(as.character(ringing_data$Wing))

ringing_data$Culmen <- as.numeric(as.character(ringing_data$Culmen)) 

ringing_data$Head <- as.numeric(as.character(ringing_data$Head))

ringing_data$Tarsus <- as.numeric(as.character(ringing_data$Tarsus))
```


A common issue in the SAFRING database is that volunteers fill in some metric data in with a zero as opposed to leaving it blank. Therefore, these zeros need to be removed as a zero length - for example - does not make sense. 

```{r step 7, results='hide'}

ringing_data$Mass[ringing_data$Mass==0] <- NA

ringing_data$Wing[ringing_data$Wing==0] <- NA

ringing_data$Head[ringing_data$Head==0] <- NA

ringing_data$Culmen[ringing_data$Culmen==0] <- NA

ringing_data$Tarsus[ringing_data$Tarsus==0] <- NA

ringing_data$Tail[ringing_data$Tail==0] <- NA

```


It is necessary - then - to remove non-sensical, extreme outliers from the metric data (measurements outside the 99% quantile). A reference set is created with the focal species and the pentad data is filtered.

```{r step 8, results='hide'}
  
checkmass <- ringing_data%>%filter(!is.na(Mass))%>%do(data.frame(lc=quantile(.$Mass, c(.05)), uq=quantile(.$Mass, c(.995)), n=length(.$Mass) ))

checkWing <- ringing_data%>%filter(!is.na(Wing))%>%do(data.frame(lcwing=quantile(.$Wing, c(.005)), uqwing=quantile(.$Wing, c(.995)), n=length(.$Wing)))

checkTarsus <- ringing_data%>%filter(!is.na(Tarsus))%>%do(data.frame(lctarsus=quantile(.$Tarsus, c(.005)), uqtarsus=quantile(.$Tarsus, c(.995)), n=length(.$Tarsus)))

checkHead <- ringing_data%>%filter(!is.na(Head))%>%do(data.frame(lcHead=quantile(.$Head, c(.005)), uqHead=quantile(.$Head, c(.995)), n=length(.$Head)))

checkCulmen <- ringing_data%>%filter(!is.na(Culmen))%>%do(data.frame(lcCulmen=quantile(.$Culmen, c(.005)), uqCulmen=quantile(.$Culmen, c(.995)), n=length(.$Culmen)))

checkTail <- ringing_data%>%filter(!is.na(Tail))%>%do(data.frame(lcTail=quantile(.$Tail, c(.005)), uqTail=quantile(.$Tail, c(.995)), n=length(.$Tail)))


ringing_data$masscheck <- NA

ringing_data$wingcheck <- NA

ringing_data$headcheck <- NA

ringing_data$culmencheck <- NA

ringing_data$tailcheck <- NA
```


We then need to validate each measurement (inside the 99 quantile) by starting a loop through the data of the focal species.

```{r step 9, results='hide'}

for(k in 1:nrow(ringing_data)){
  
  ringing_data$masscheck[k] <- ifelse(ringing_data$Mass[k]<checkmass$uq & ringing_data$Mass[k]>checkmass$lc , "OK", "error")  
  
  ringing_data$wingcheck[k] <- ifelse(ringing_data$Wing[k]<checkWing$uq & ringing_data$Wing[k]>checkWing$lc , "OK", "error") 
  
  ringing_data$tarsuscheck[k] <- ifelse(ringing_data$Tarsus[k]<checkTarsus$uq & ringing_data$Tarsus[k]>checkTarsus$lc , "OK", "error")  
  
  ringing_data$headcheck[k] <- ifelse(ringing_data$Head[k]<checkHead$uq & ringing_data$Head[k]>checkHead$lc , "OK", "error") 
  
  ringing_data$culmencheck[k] <- ifelse(ringing_data$Culmen[k]<checkCulmen$uq & ringing_data$Culmen[k]>checkCulmen$lc , "OK", "error") 
  
  ringing_data$tailcheck[k] <- ifelse(ringing_data$Tail[k]<checkTail$uq & ringing_data$Tail[k]>checkTail$lc , "OK", "error")  
}  
```


The 'NAs' then need to be replaced with 'OK' to soften the next filter step.

```{r step 10, results='hide'}

ringing_data$masscheck[is.na(ringing_data$masscheck)] <- "OK"

ringing_data$tarsuscheck[is.na(ringing_data$tarsuscheck)] <- "OK"

ringing_data$headcheck[is.na(ringing_data$headcheck)] <- "OK"

ringing_data$culmencheck[is.na(ringing_data$culmencheck)] <- "OK"

ringing_data$tailcheck[is.na(ringing_data$tailcheck)] <- "OK"
```


The object containing the data for the focal species then needs to be redefined with the error filtered out. 

```{r step 11, results='hide'}

ringing_data <- filter(ringing_data, masscheck!='error', tarsuscheck!='error',headcheck!='error',culmencheck!='error',tailcheck!='error')

```

The column containing the brood patch information then needs to be cleaned so that the correct, and tidied data is plotted in Section 3. 
```{r step 12, results='hide', warning=FALSE}

ringing_data$BP <-  as.numeric(ringing_data$Brood_patch) 

ringing_data$BP <-  ifelse(ringing_data$BP >0 , 1, ringing_data$BP)

ringing_data$BP <-  ifelse(ringing_data$BP <0 , 0, ringing_data$BP)

ringing_data$BP <-  ifelse(ringing_data$Brood_patch %in% c("1.0", "2.0", "y", "BP", "Brood Patch") , 1, ringing_data$BP)

ringing_data$BP <-  ifelse(ringing_data$Brood_patch %in% c("-1.0", "n", "N", "no") , 0, ringing_data$BP)

```


The date information that is recorded in the SABAP2 database has to be cleaned and organized as well. 
```{r step 13, results='hide'}

ringing_data$Startdate1 <- as.Date(ringing_data$Startdate, "%Y-%m-%d")

ringing_data$Month <- format(ringing_data$Startdate1, "%m") 

ringing_data$Month <- as.numeric(ringing_data$Month)

ringing_data$Year <- format(ringing_data$Startdate1, "%Y") 

```


Next, the moult data must be numeric and filtered to remove NA values, which imply that active moult is not occurring. 
```{r step 14, results='hide', warning=FALSE}

ringing_data$Moult <- as.numeric(ringing_data$Moult)

ringing_data <- filter(ringing_data, !is.na(Moult))

ringing_data$gender <- ifelse(ringing_data$Sex==1, "Male", ifelse(ringing_data$Sex==2, "Female", NA))

ringing_data$active_moult <- ifelse(ringing_data$Moult>0&ringing_data$Moult<5555555555, 1, 0)
```


### Section 3

The geographic pentads in the SAFRING database contain spatial information (latitude and longitude). However, to work with this spatial information, latitude and longitude values contained within the pentad object need to be separated to tidy the data and work with the "sf" package. 

The functions that we need to get sections of the pentads for mapping.
```{r,step 15, results='hide'}

left = function(text, num_char) {substr(text, 1, num_char)}
mid = function(text, start_num, num_char) {  substr(text, start_num, start_num + num_char - 1)}
```

We then have to tidy the pentad data to pull out rough latitude and longitude values. 
```{r, step 16, results='hide'}

ringing_data$Lat <- as.numeric(left(ringing_data$pentad, 2)) +  as.numeric(mid(ringing_data$pentad, 3, 2))/60
ringing_data$Latitude <- ifelse(mid(ringing_data$pentad,5,1)=="_", ringing_data$Lat*-1, ringing_data$Lat)
ringing_data$Longitude <- as.numeric(mid(ringing_data$pentad, 6,2)) +  as.numeric(mid(ringing_data$pentad, 8, 2))/60
```

We now have to objects called Latitude and Longitude in the ringing data for the malachite sunbird so that we can map the localities of this species across South Africa. This brings us to Section 4. 


### Section 4

Firstly, we have to coarsely split the spatial data by longitude. For the purposes of the map that we wish to produce for the project, we do not need exact longitude and latitude values. We need to split the values of latitude and longitude into 'east' and 'west' because we want to visualize which localities of the malachite sunbird - or another focal species that stradles both regions - are in the 'east' and which are in the 'west'. 

```{r step 17, results='hide'}

ringing_data$West_East <- ifelse(ringing_data$Longitude<27, "West", "East")
```

Let's check the class of the variable containing the spatial data. 
```{r step 18}
class(ringing_data$West_East)
```

We can see that this variable is a 'character'. Therefore, we need to change this variable to a data frame. 

```{r step 19, results='hide'}
Locations <- as.data.frame(ringing_data$West_East)
```


```{r}
class(ringing_data)
```


```{r step 20}
class(Locations)
```
There we go! That is better. 

Now, this data frame has to be converted to a spatial object. We have to convert this object 'Locations' of class = sf. The coordinate reference system (CRS) also have to be defined. The CRS should be defined to be Geographic WGS84 using the EPSG code (4326).  

```{r step 21}

Locations1 <- st_as_sf(ringing_data, coords = c("Longitude", "Latitude"), crs = 4326)
```

```{r}
class(Locations1)
```
```{r}
names(Locations1)
```
```{r}

ggplot(ringing_data, aes(Longitude, Latitude, colour = West_East))+geom_point()
```

```{r}
install.packages("rosm", repos = "http://cran.us.r-project.org")
```

```{r}
install.packages("ggspatial", repos = "http://cran.us.r-project.org")
```


```{r}
library(rosm)
library(ggspatial)
```


```{r}
ggplot() + 
  annotation_map_tile(type = "osm", progress = "none", zoomin = -1) + 
  geom_sf(data=Locations1, aes(colour = West_East))
```
