---
title: "GIS Code"
author: "Rebecca Irons (IRNREB001)"
date: '2022-04-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GIS Assignment

#### Section 1: Install and Load Packages 
#### Section 2: Tidy SAFRING Data
#### Section 3: Tidy Geographic Pentad Data
#### Section 4: Creating the Map 

This document provides step-by-step instructions to create a map of the localities of where 
malachite sunbirds have been ringed across South Africa, from east by KwaZulu-Natal to west. The data for this species is accessed from SAFRING using an API. For more details on the SAFRING database, see <https://safring.birdmap.africa/>. 

### Section 1

In this section, we will install and load the necessary packages to read in the data from the SAFRING database, tidy the data and use GIS tools to create a map. 

First, the workspace must be clean. 
```{r step 1}
rm(list = ls())
```

Install the necessary packages: 

- Tidyverse is required to manipulate the data: 
```{r step 2a, results='hide', warning=FALSE}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
```

- RCurl and rjson is required to access the data from SABAP2 using as API.  
```{r step 2b, results='hide', warning=FALSE}
install.packages("RCurl", repos = "http://cran.us.r-project.org")

install.packages("rjson", repos = "http://cran.us.r-project.org")
```

- sf is required to work with vector data to create a map
```{r step 2c, results='hide', warning=FALSE}

install.packages("sf", repos = "http://cran.us.r-project.org")

```


The packages from the tidyverse have to be loaded. These packages are:

```{r step 3a, results='hide', warning=FALSE, message=FALSE}

library(dplyr)

library(ggplot2)

library(broom)

```


The packages required for loading in the data from the SAFRING website are:
```{r step 3 b, results='hide',warning=FALSE, message=FALSE}

library(RCurl)

library(rjson)
```

The packages required for working with the vector data are: 
```{r step 3c, results='hide', warning=FALSE, message=FALSE}

library(sf)
```


Using an API, access the SAFRING database and load the data for your focal species. 

```{r step 4, results='hide'}
SPP = 751
ringing_data <- read.csv(paste('https://api.birdmap.africa/safring/species/records_list/',SPP,'?format=csv',sep=""))
```


### Section 2



